String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-10-13 2023-1-19/sections/introduction.tex
review 2

Regular expressions (regex for short) and string-length function are widely used in string-manipulating programs. According to the statistics from \cite{CS16,DCSL18,WS18}, regular expressions are used in about 30–40\% of Java, JavaScript, and Python software projects. 
Moreover, string-length occupies 78\% of the occurrences of string operations in 18 Javascript applications, according to the statistics from \cite{malware_detection_3_kudzu}. 
%Operations like \verb|length| that take string inputs and return an integer frequently appear in real-world JavaScript applications (78\% of string operations in 18 applications)
As a result, most of the aforementioned string constraint solvers support both regular expressions and string-length function. Moreover, specialized algorithms have been proposed to solve such string constraints efficiently. \cite{LTR+15,BD+23}. 
%a decision procedure based on tableaux calculus was designed in \cite{LTR+15} for the class of string constraints with regular expressions and string-length function. The decision procedure has been implemented in CVC4/5 \cite{LRT+16}.

Counting (aka repetition) is a convenient feature in regular expressions that count the number of matchings of sub-patterns. For instance, $a^{\{2, 4\}}$ specifies that $a$ occurs at least twice and at most four times, and $a^{\{2, \infty\}}$ specifies that $a$ occurs at least twice. 
Note that the Kleene star and the Kleene plus operator are special cases of counting. For instance, $a^*$ are equivalent to $a^{\{0,\infty\}}$ and $a^+$ are equivalent to $a^{\{1,\infty\}}$.
Counting is a frequently used feature of regular expressions. According to the statistics from \cite{CS16}, Kleene star/plus occur in more than 70\% of 1,204 Python projects, while other forms of counting occur in more than 20\% of them. Therefore, an efficient analysis of string manipulating programs requires efficient solvers for string constraints containing regular expressions with counting\footnote{In the rest of this paper, for clarity, we use counting to denote expressions of the form $e^{\{m, n\}}$ and  $e^{\{m, \infty\}}$, but not $e^*$ or $e^+$.} and string-length function at least.

>>>>

Regular expressions (regex for short) and the string-length function are widely used in string-manipulating programs. According to the statistics from \cite{CS16,DCSL18,WS18}, regular expressions are used in about 30–40\% of Java, JavaScript, and Python software projects. 
Moreover, string-length occupies 78\% of the occurrences of string operations in 18 Javascript applications, according to the statistics from \cite{malware_detection_3_kudzu}. 
%Operations like \verb|length| that take string inputs and return an integer frequently appear in real-world JavaScript applications (78\% of string operations in 18 applications)
As a result, most of the aforementioned string constraint solvers support both regular expressions and string-length function. Moreover, specialized algorithms have been proposed to solve such string constraints efficiently. \cite{LTR+15,BD+23}. 
%a decision procedure based on tableaux calculus was designed in \cite{LTR+15} for the class of string constraints with regular expressions and string-length function. The decision procedure has been implemented in CVC4/5 \cite{LRT+16}.

Counting (aka repetition) is a convenient feature in regular expressions that count the number of matchings of sub-patterns. For instance, $a^{\{2, 4\}}$ specifies that $a$ occurs at least twice and at most four times, and $a^{\{2, \infty\}}$ specifies that $a$ occurs at least twice. 
Note that the Kleene star and the Kleene plus operator are special cases of counting. For instance, $a^*$ is equivalent to $a^{\{0,\infty\}}$ and $a^+$ are equivalent to $a^{\{1,\infty\}}$.
Counting is a frequently used feature of regular expressions. According to the statistics from \cite{CS16}, Kleene star/plus occur in more than 70\% of 1,204 Python projects, while other forms of counting occur in more than 20\% of them. Therefore, an efficient analysis of string manipulating programs requires efficient solvers for string constraints containing regular expressions with counting\footnote{In the rest of this paper, for clarity, we use counting to denote expressions of the form $e^{\{m, n\}}$ and  $e^{\{m, \infty\}}$, but not $e^*$ or $e^+$.} and string-length function at least.

### 2023-10-13 2023-1-19/sections/overview.tex
latest version
review 2

First, we construct an CEFA for the regular expression $(\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$. Three registers are introduced, say $r_1, r_2, r_3$, to represent the three counting operators;  
%Moreover, we introduce another register, say $r_0$, to denote the length of strings. 
the nondeterministic finite automaton (NFA) for $(\Sigma \setminus a)^* (\Sigma \setminus b)^* (\Sigma \setminus c)^*$ is constructed; the updates of registers is added to the transitions of the NFA; the counting bounds are specified by the accepting condition $1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60$, resulting in a CEFA $\aut_1$ illustrated in Figure~\ref{fig:overview}(a). $r_1++$ means that we increment the value of $r_1$ by one after running the transition.
%Then we obtain a cost-enriched finite automaton (CEFA) $\aut_1$ that is illustrated in Figure~\ref{fig:overview}(a), where the self-loops around $p_0, p_1, p_2$ represent the first, second, and third counting operator respectively, in which $r_1, r_2, r_3$ are incremented. 
%Moreover, $r_0$ is incremented in all the three self-loops. 
A string $w$ is accepted by $\aut_1$ if, when reading the characters in $w$, $\aut_1$ applies the transitions to update the state and the values of registers, reaching a final state $q$ in the end, and the resulting values of the three registers, say $v_1,v_2,v_3$, satisfy the accepting condition. In addition, we construct other two CEFAs $\aut_2$ for $\Sigma^* c^+$ (see Figure~\ref{fig:overview}(b)) and $\aut_3$ for string length function (see Figure~\ref{fig:overview}(c)). In $\aut_3$, a register $r_4$ is used to denote the length of strings and the accepting condition is $\ltrue$ (See Section~\ref{subsec:regex2cefa} for more details about the construction of CEFA.) Note that we represent the counting operators symbolically by registers instead of unfolding them explicitly. 
%where the register $r_0$ is used to record the length of strings.

@@

Nevertheless, when the original regexes are complicated (e.g. contains the occurrences of negation or intersection operators), the sizes of the NFA $\cB$ can still be big and the sizes of the LIA formulas defining the Parikh image of $\cB$ are also big. Since the satisfiability of LIA formulas is an NP-complete problem \cite{Haase18}, big sizes of LIA formulas would be a bottleneck of the efficiency. 
To tackle this issue, we propose techniques to reduce the sizes of the NFA $\cB$.
%over the alphabet $\costset$.

>>>>

First, we construct an CEFA for the regular expression $(\Sigma \setminus a)^{\{1, 60\}} (\Sigma \setminus b)^{\{1, 60\}} (\Sigma \setminus c)^{\{0, 60\}}$. Three registers are introduced, say $r_1, r_2, r_3$, to represent the three counting operators;  
%Moreover, we introduce another register, say $r_0$, to denote the length of strings. 
the nondeterministic finite automaton (NFA) for $(\Sigma \setminus a)^* (\Sigma \setminus b)^* (\Sigma \setminus c)^*$ is constructed; the updates of registers are added to the transitions of the NFA; the counting bounds are specified by the accepting condition $1 \le r_1 \le 60 \wedge 1 \le r_2 \le 60 \wedge 0 \le r_3 \le 60$, resulting in a CEFA $\aut_1$ illustrated in Figure~\ref{fig:overview}(a). $r_1++$ means that we increment the value of $r_1$ by one after running the transition.
%Then we obtain a cost-enriched finite automaton (CEFA) $\aut_1$ that is illustrated in Figure~\ref{fig:overview}(a), where the self-loops around $p_0, p_1, p_2$ represent the first, second, and third counting operator respectively, in which $r_1, r_2, r_3$ are incremented. 
%Moreover, $r_0$ is incremented in all the three self-loops. 
A string $w$ is accepted by $\aut_1$ if, when reading the characters in $w$, $\aut_1$ applies the transitions to update the state and the values of registers, reaching a final state $q$ in the end, and the resulting values of the three registers, say $v_1,v_2,v_3$, satisfy the accepting condition. In addition, we construct other two CEFAs $\aut_2$ for $\Sigma^* c^+$ (see Figure~\ref{fig:overview}(b)) and $\aut_3$ for string length function (see Figure~\ref{fig:overview}(c)). In $\aut_3$, a register $r_4$ is used to denote the length of strings and the accepting condition is $\ltrue$ (See Section~\ref{subsec:regex2cefa} for more details about the construction of CEFA.) Note that we represent the counting operators symbolically by registers instead of unfolding them explicitly. 
%where the register $r_0$ is used to record the length of strings.

@@

Nevertheless, when the original regexes are complicated (e.g. contain occurrences of negation or intersection operators), the sizes of the NFA $\cB$ can still be big and the sizes of the LIA formulas defining the Parikh image of $\cB$ are also big. Since the satisfiability of LIA formulas is an NP-complete problem \cite{Haase18}, big sizes of LIA formulas would be a bottleneck of the performance. 
To tackle this issue, we propose techniques to reduce the sizes of the NFA $\cB$.
%over the alphabet $\costset$.
